---
title: "SQL project"
---

**NOTE: This demonstration has been adapted from Jo Hardin's [classnotes](https://st47s.com/SDS261/Notes/).**

**NOTE: This is joint work with Abhishek Chakraborty.**

# load packages and connect to database server

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
library(duckdb)   # R package for building a duckdb database locally
```


As an example, we will consider the Saturday Night Live datasets available on the *snldb* GitHub [repo](https://github.com/hhllcks/snldb). Data is scraped from [SNL archives](http://www.snlarchives.net) and [IMDb](http://www.imdb.com/title/tt0072562) by Hendrik Hilleckes and Colin Morris.

```{r}

con_snl <- DBI::dbConnect(duckdb::duckdb(), dbdir = "snl_db")

```


```{sql connection=con_snl}
SHOW TABLES;
```


# accessing data (csv files) from GitHub (or, elsewhere)

Notice that there are eleven *.csv* files available in the *output* folder. Let's look at the *casts* dataset in R, so that we understand the data we want to load into the database.

```{r}
# read in the data

casts <- readr::read_csv("https://raw.githubusercontent.com/hhllcks/snldb/refs/heads/master/output/casts.csv")

```


```{r}

glimpse(casts)

```


```{r}
summary(casts)

```


```{r}

# save the dataset as a .csv file on your (local) computer

write_csv(casts, "casts.csv")   

```


# data types in SQL

Consult [data types](https://dev.mysql.com/doc/refman/8.0/en/data-types.html) for details.


# importing data

Once the database is set up, you will be ready to import .csv files into the database as tables. Importing .csv files as tables requires a series of steps:

* a *USE* statement that ensures we are in the right database.
* a series of *DROP TABLE* statements that drop any old tables with the same names as the ones we are going to create.
* a series of *CREATE TABLE* statements that specify the table structures and field types.
* a series of *COPY* (or *LOAD DATA* for MySQL) statements that read the data from the .csv files into the appropriate tables.


```{sql connection=con_snl}

USE snl_db;

```


```{sql connection=con_snl}

DROP TABLE IF EXISTS casts;

```


```{sql connection=con_snl}


CREATE TABLE casts (
  aid VARCHAR(255) NOT NULL DEFAULT ' ',
  sid SMALLINT NOT NULL DEFAULT 0,
  featured BOOLEAN NOT NULL DEFAULT FALSE,  /* use TINYINT(1) in MySQL instead of BOOLEAN) */
  first_epid INTEGER DEFAULT NULL,
  last_epid INTEGER DEFAULT NULL,
  update_anchor BOOLEAN NOT NULL DEFAULT FALSE,
  n_episodes SMALLINT NOT NULL DEFAULT 0,
  season_fraction DOUBLE NOT NULL DEFAULT 0,
  PRIMARY KEY (sid, aid)
);

```


```{sql connection=con_snl}

COPY casts FROM 'casts.csv' (FORMAT CSV, HEADER, NULL 'NA');

```


# checks

```{sql connection=con_snl}
SHOW DATABASES;
```


```{sql connection=con_snl}
SHOW TABLES;
```


```{sql connection=con_snl}
DESCRIBE casts;
```


```{sql connection=con_snl}
SELECT * FROM casts LIMIT 10;
```


# data visualization

Let's try to create a simple visualization here.

```{sql connection=con_snl}

-- make sure to check the total number of observations before collecting the dataset in local Environment

SELECT COUNT(*) AS num_obs
FROM casts;

```


```{r}

# save dataset in local Environment

casts_df <- casts |>  dplyr::collect()   


```


```{r}

casts_df |> 
  count(aid) |> 
  arrange(desc(n)) |> 
  slice_head(n = 10) |> 
  ggplot() +
  geom_col(aes(x = fct_reorder(aid, n, .desc = TRUE), y = n),
           color = "white", fill = "lightblue") +
  labs(x = "Actors", y = "Number of episodes",
       title = "Top Ten Appearances") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```


# terminate the SQL connection

```{r}

DBI::dbDisconnect(con_snl, shutdown = TRUE)

```


