---
title: "text analysis project"
---

**NOTE: This demonstration has been adapted from the [MDSR (3e)](https://mdsr-book.github.io/mdsr3e/) textbook.**

**NOTE: This is joint work with Abhishek Chakraborty.**


# resources

-   [arXiv.org](https://arxiv.org/)


# packages

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(scales)
library(tidytext)
library(textdata)
library(mdsr)
library(stopwords)
library(slam)
library(tm)
```


# data

We can import data using the *aRxiv* package. We won't run it here since it strains the arXiv server.

```{r}
#| eval: false

library(aRxiv)

df <- arxiv_search(
  query = 'ti:"Data Science"',
  limit = 20000,
  batchsize = 100
)

# glimpse(df)

```

Instead, for this case study we will use some already-collected data from the *mdsr* package.

```{r}

library(mdsr) 

data(DataSciencePapers)

?DataSciencePapers

glimpse(DataSciencePapers)

```

### initial data exploration and cleaning

```{r}

DataSciencePapers <- DataSciencePapers |>
  mutate(
    submitted = lubridate::ymd_hms(submitted), 
    updated = lubridate::ymd_hms(updated)
  )

glimpse(DataSciencePapers)

```

```{r}
#| message: false
#| warning: false

mosaic::tally(~ year(submitted), data = DataSciencePapers)

```

```{r}

DataSciencePapers |>
  filter(id == "1809.02408v2") |>
  glimpse()

```

```{r}

DataSciencePapers |>
  group_by(primary_category) |>
  count()

```

```{r}

DataSciencePapers <- DataSciencePapers |>
  mutate(field = str_extract(primary_category, pattern = "^[a-z,-]+"))

mosaic::tally(x = ~field, margins = TRUE, data = DataSciencePapers) |>
  sort()

```

### corpora and tokenization

```{r}

DataSciencePapers |>
  unnest_tokens(output = word, input = abstract, token = "words") |>
  count(id, word, sort = TRUE)

```

```{r}

arxiv_words <- DataSciencePapers |>
  unnest_tokens(output = word, input = abstract, token = "words") |>
  anti_join(tidytext::get_stopwords(), by = "word")

arxiv_words |>
  count(id, word, sort = TRUE)

```

```{r}

arxiv_abstracts <- arxiv_words |>
  group_by(id) |>
  summarize(abstract_clean = paste(word, collapse = " "))

arxiv_papers <- DataSciencePapers |>
  left_join(arxiv_abstracts, by = "id")

```

### wordcloud

```{r}
#| message: false
#| warning: false

library(wordcloud)

set.seed(4302026)
arxiv_papers |>
  pull(abstract_clean) |>
  wordcloud(
    max.words = 40, 
    scale = c(8, 1), 
    colors = topo.colors(n = 30), 
    random.color = TRUE
  )

```


### sentiment analysis

We will use the [AFINN](https://arxiv.org/abs/1103.2903) lexicon, where each word is associated with an integer value  ranging from -5 to 5, for our sentiment analysis purposes. When prompted, make sure to download the lexicon.

```{r}
#| message: false
#| warning: false

afinn <- tidytext::get_sentiments("afinn")

head(afinn)

```


We can join this lexicon with our data to calculate a sentiment score.


```{r}

arxiv_words |> 
  inner_join(afinn, by = "word") |>
  select(word, id, value)

```

```{r}

arxiv_sentiments <- arxiv_words |> 
  left_join(afinn, by = "word") |>
  group_by(id) |>
  summarize(num_words = n(),
    sentiment = sum(value, na.rm = TRUE), 
    .groups = "drop") |>
  mutate(sentiment_per_word = sentiment / num_words) |>
  arrange(desc(sentiment))

arxiv_papers <- arxiv_papers |>
  left_join(arxiv_sentiments, by = "id")

arxiv_papers |>
  skim(sentiment, sentiment_per_word)

```

The average sentiment score of these papers is 4, but they range from -26 to 39. Since abstracts with more words might accrue a higher sentiment score, we control for abstract length by dividing by the number of words. The paper with the highest sentiment score per word had a score of 0.333. Let's take a closer look at the most positive abstract.

```{r}

most_positive <- arxiv_papers |>
  filter(sentiment_per_word == max(sentiment_per_word)) |>
  pull(abstract)

strwrap(most_positive)

```

Let's explore if there are time trends or differences between different disciplines.

```{r}
#| message: false
#| warning: false

ggplot(data = arxiv_papers, 
  aes(x = submitted, y = sentiment_per_word, 
    color = field == "cs")) + 
  geom_smooth(se = TRUE) + 
  scale_color_brewer("Computer Science?", palette = "Set2") +
  labs(x = "Date submitted", y = "Sentiment score per word")

```

 
### bigrams

```{r}

arxiv_bigrams <- arxiv_papers |>
  unnest_tokens(
    output = arxiv_bigram, 
    input = abstract_clean, 
    token = "ngrams", 
    n = 2
  ) |>
  dplyr::select(arxiv_bigram, id)

arxiv_bigrams

arxiv_bigrams |>
  count(arxiv_bigram, sort = TRUE)

```

### tf-idf and document term matrices

```{r}

arxiv_words |>
  count(word) |>
  arrange(desc(n)) |>
  head()

```

```{r}

tidy_DTM <- arxiv_words |>
  count(id, word) |>
  tidytext::bind_tf_idf(word, id, n)

tidy_DTM

```

```{r}

tidy_DTM |>
  arrange(desc(tf)) |>
  head()

```

```{r}

tidy_DTM |>
  arrange(desc(idf), desc(n)) |>
  head()

```

```{r}

tidy_DTM |>
  filter(word == "wildfire")

```

```{r}

arxiv_papers |>
  pull(abstract) |>
  str_subset("wildfire") |>
  strwrap() |>
  head()

```

```{r}

tidy_DTM |>
  filter(word == "implications")

```

```{r}

tidy_DTM |>
  filter(id == "1809.02408v2") |>
  arrange(desc(tf_idf)) |>
  head()

```

```{r}

tm_DTM <- arxiv_words |>
  count(id, word) |>
  tidytext::cast_dtm(document = id, term = word, value = n)

tm_DTM

```

